---
title: "Assignment 10: Data Scraping"
author: "Jack Killeen"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE, warning=FALSE}
#1 
library(tidyverse)
library(rvest)
library(janitor)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
LWSP <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
LWSP_Durham <- read_html(LWSP)

all_tables <- LWSP_Durham %>% html_table(fill = TRUE)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
all_tables <- LWSP_Durham %>% html_table(fill = TRUE)

durham_info <- all_tables[[2]]
water_system_name <- durham_info[1, 2]
pwsid <- durham_info[1, 5]
ownership <- durham_info[2, 5]

max_day_use_table <- all_tables[[13]]

max_day_use <- c(
  max_day_use_table[[3]],
  max_day_use_table[[6]],
  max_day_use_table[[9]]
) %>% as.numeric()

water_system_name
pwsid
ownership
max_day_use

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data, warning=FALSE}
#4 
scraped_months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

durham_df <- tibble(
  Month = factor(scraped_months, levels = month.abb),
  Date = as.Date(paste0("2024-", match(scraped_months, month.abb), "-01")),
  Max_Day_Use_MGD = max_day_use,
  Water_System_Name = rep(water_system_name, 12),
  PWSID = rep(pwsid, 12),
  Ownership = rep(ownership, 12)
)

print(durham_df)

#5 
ggplot(durham_df, aes(x = Month, y = Max_Day_Use_MGD, group = 1)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "2024 Maximum Daily Water Withdrawal by Month",
    x = "Month",
    y = "Max Day Use (MGD)"
  ) +
  theme_minimal()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two input - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
get_lwsp_data <- function(pwsid, year) {
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", pwsid, "&year=", year)
  page <- read_html(url)
  all_tables <- page %>% html_table(fill = TRUE)
  sys_info <- all_tables[[2]]
  water_system_name <- sys_info[1, 2]
  ownership <- sys_info[2, 5]
  max_day_use_table <- all_tables[[13]]
  max_day_use <- c(
    max_day_use_table[[3]],
    max_day_use_table[[6]],
    max_day_use_table[[9]]
  ) %>% as.numeric()
  months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  df <- tibble(
    Month = factor(months, levels = month.abb),
    Date = as.Date(paste0(year, "-", match(months, month.abb), "-01")),
    Max_Day_Use_MGD = max_day_use,
    Water_System_Name = rep(water_system_name, 12),
    PWSID = rep(pwsid, 12),
    Ownership = rep(ownership, 12),
    Year = rep(year, 12)
  )
  return(df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020

```{r fetch.and.plot.Durham.2020.data}
#7 
durham_2020 <- get_lwsp_data("03-32-010", 2020)
print(durham_2020)

ggplot(durham_2020, aes(x = Month, y = Max_Day_Use_MGD, group = 1)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Durham 2020 Maximum Daily Water Usage by Month",
    x = "Month",
    y = "Max Day Use"
  ) +
  theme_minimal()

```

8. Use the function above to extract data for Asheville (PWSID = '01-11-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2020.data}
#8 
asheville_2020 <- get_lwsp_data("01-11-010", 2020)

combined_df <- bind_rows(durham_2020, asheville_2020)

combined_df <- combined_df %>%
  mutate(City = ifelse(PWSID == "03-32-010", "Durham", "Asheville"))

print(combined_df)

ggplot(combined_df, aes(x = Month, y = Max_Day_Use_MGD, color = City, group = City)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Comparison of 2020 Maximum Daily Water Withdrawals",
    x = "Month",
    y = "Max Day Use",
    color = "City"
  ) +
  theme_minimal()

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
years <- 2018:2023
pwsid <- rep("01-11-010", length(years))

asheville_all_years <- map2_df(pwsid, years, get_lwsp_data)
glimpse(asheville_all_years)

ggplot(asheville_all_years, aes(x = Month, y = Max_Day_Use_MGD, group = Year, 
                                color = as.factor(Year))) +
  geom_line(size = 1) +
  geom_point(size = 1.8) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black", 
              linetype = "dashed", size = 1) +
  labs(
    title = "Asheville 2018 to 2023 Max Daily Water Withdrawals",
    x = "Month",
    y = "Max Day Use",
    color = "Year"
  ) +
  theme_minimal()

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Asheville has a trend in water usage over time with the winter months (Dec/Jan) having a spike, before a slow down in the fall and spring, but another spike in the summer.
>
